Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-13 paper.tex
some more updates

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking, pharmacophore modeling and proteochemometric methods
have been designed to take into account both ligand and receptor
structures. The last method is an example of an extension of
statistical QSAR methods to simultaneous model the f receptor and
ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.

@@

In order to judge which predictive modeling approach is applicable in
every case, we firstly have to consider the goal of a the model in a
cheminformatics setting. In some cases, pure predictive ability is
desired -- such as in a virtual screening setting. (Basically, only
the reliability of the model matters, but other aspects such as
interpretability are of minor importance.) For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate better-predicted
\emph{in silico} compounds into real-world compounds. In such cases,
``black-box methods'' can be less useful, though attempts have been
made to interpret vectorial models built using neural networks,
na\"{i}ve Bayes, Support Vector Machines or random forests as well.

@@

With the advent of high-throughput screening technologies, large
libraries of compounds can be screened against multiple targets in an
efficient manner. Such panel assays provide a broad, systems-level
view of small molecule activities.  Models developed on such data
afford us the opportunity to identify targets, characterize off-target
effects and so on. However, most approaches to this problem are rather
simplistic since they develop multiple individual models
\cite{Chen:2010zr}, leading to multiple, independent predictions for a
given input molecule.  Instead, one could imagine an approach that
takes into account the covariance structure of the multiple observed
activities and the structural descriptors within a single model. Such
an approach could lead to more robust predictions for panel
assays. Finally, this approach also better reflects clinically
relevant compound profiles~\cite{kuhn2010} and a ``personalized
medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).

@@

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been
noted, recently large structure and bioactivity data collections have
become available. As a result, this enables much more robust
validation of cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques that are employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open'' in nature. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and there have been efforts to mine
these for chemical information using Open tools.

>>>>

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking, pharmacophore modeling and proteochemometric methods
have been designed to take into account both ligand and receptor
structures. The last method is an example of an extension of
statistical QSAR methods to simultaneous model the receptor and
ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.

@@

In order to judge which predictive modeling approach is applicable in
every case, we firstly have to consider the goal of a the model in a
cheminformatics setting. In some cases, pure predictive ability is
desired -- such as in a virtual screening setting. For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate the predicted
outputs into real-world compounds. In such cases, ``black-box
methods'' can be less useful, though attempts have been made to
interpret vectorial models built using neural networks, na\"{i}ve
Bayes, Support Vector Machines or random forests.

@@

With the advent of high-throughput screening technologies, large
libraries of compounds can be screened against multiple targets in an
efficient manner. Such panel assays provide a broad, systems-level
view of small molecule activities.  Models developed on such data
afford us the opportunity to identify targets, characterize off-target
effects and so on. However, most approaches to this problem temd tp
develop multiple individual models \cite{Chen:2010zr}, leading to
multiple, independent predictions for a given input molecule.
Instead, one could imagine an approach that takes into account the
covariance structure of the multiple observed activities and the
structural descriptors within a single model. Such an approach could
lead to more robust predictions for panel assays. Finally, this
approach also better reflects clinically relevant compound
profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
(\emph{e.g.}, drug-drug interaction profiles).

@@

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been
noted, recently large structure and bioactivity data collections have
become available. As a result, this enables much more robust
validation of cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques that are employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open'' in nature. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and represents an excellent resource
to support these types of efforts.
